---
output: html_document
editor_options: 
  chunk_output_type: console
---
Lesson 5
========================================================
```{r}
library(ggplot2)
library(tidyr)
library(dplyr)
library(gtools)
library(gridExtra)
library(tibble)
library(data.table)

pf <- read.csv('pseudo_facebook.tsv', sep='\t')

```

### Multivariate Data
Notes:

***

### Moira Perceived Audience Size Colored by Age
Notes:

***

### Third Qualitative Variable
Notes:

```{r Third Qualitative Variable}
#7.3: Show mean as an 'x' (shape = 4)
ggplot(aes(x = gender, y = age),data = subset(pf, !is.na(gender))) + 
  geom_boxplot() + 
  stat_summary(fun.y=mean, geom = 'point', shape=4)

#: view gender mean as line chart
ggplot(aes(x = age, y = friend_count),data = subset(pf, !is.na(gender))) + 
  geom_line(aes(color=gender), stat = 'summary', fun.y=median)


#Quiz - group by 2 variables
pf.fc_by_age_gender <- pf %>%
  group_by(age, gender) %>%
  filter(!is.na(gender)) %>%
  summarize(mean_friend_count = mean(friend_count),
            median_friend_count = median(friend_count),
            n=n()) %>%
  ungroup() %>%
  arrange(age, gender)

#7.4: plot using summarized data
ggplot(aes(x=age, y=median_friend_count), data = pf.fc_by_age_gender) +
  geom_line(aes(color=gender))

```


***

### Plotting Conditional Summaries
Notes:

```{r Plotting Conditional Summaries}

```

***

### Thinking in Ratios
Notes:

***

### Wide and Long Format
Notes:

***

### Reshaping Data
Notes:

```{r}

library(reshape2)

pf.fc_by_age_gender.wide <- dcast(pf.fc_by_age_gender, 
                                 age ~ gender,
                                 value.var = 'median_friend_count')
head(pf.fc_by_age_gender.wide)

#This works!
x <- melt(pf.fc_by_age_gender.wide, id.vars = 'age', measure.vars = c("female", "male"), variable.name = 'gender', value.name = 'median_friend_count')

```


***

### Ratio Plot
Notes:

```{r Ratio Plot}

#7.8 Ratio PLot
ggplot(aes(x=age, y = female/male), data = pf.fc_by_age_gender.wide) +
  geom_line() +
  geom_hline(yintercept = 1, linetype=2, alpha = 0.3)


```

***

### Third Quantitative Variable
Notes:

```{r Third Quantitative Variable}
#7.9
pf$year_joined <- 2014 - ceiling(pf$tenure/365)


```

***

### Cut a Variable
Notes:

```{r Cut a Variable}

#7.10

# Create a new variable in the data frame
# called year_joined.bucket by using
# the cut function on the variable year_joined.

# You need to create the following buckets for the
# new variable, year_joined.bucket

#        (2004, 2009]
#        (2009, 2011]
#        (2011, 2012]
#        (2012, 2014]

pf$year_joined.bucket <- cut(pf$year_joined, breaks=c(2004,2009,2011,2012,2014))

```

***

### Plotting it All Together
Notes:

```{r Plotting it All Together}
#7.11
ggplot(aes(x = age, y = friend_count),data = subset(pf, !is.na(year_joined.bucket))) + 
  geom_line(aes(color=year_joined.bucket), stat = 'summary', fun.y=median)

```

***

### Plot the Grand Mean
Notes:

```{r Plot the Grand Mean}
#7.12
ggplot(aes(x = age, y = friend_count),data = subset(pf, !is.na(year_joined.bucket))) + 
  geom_line(aes(color=year_joined.bucket), stat = 'summary', fun.y=mean) +
  geom_line(stat='summary', fun.y = mean, linetype = 2)

```

***

### Friending Rate
Notes:

```{r Friending Rate}
#7.13

summary(with(subset(pf,tenure > 0), friend_count / tenure)) 
```

***

### Friendships Initiated
Notes:

What is the median friend rate?

What is the maximum friend rate?

```{r Friendships Initiated}
#7.14
ggplot(aes(x = tenure, y = friendships_initiated/tenure),data = subset(pf, tenure>0)) + 
  geom_line(aes(color=year_joined.bucket), stat = 'summary', fun.y=mean) 



```

***

### Bias-Variance Tradeoff Revisited
Notes:

```{r Bias-Variance Tradeoff Revisited}
#7.15

ggplot(aes(x = tenure, y = friendships_initiated / tenure),
       data = subset(pf, tenure >= 1)) +
  geom_line(aes(color = year_joined.bucket),
            stat = 'summary',
            fun.y = mean)

ggplot(aes(x = 7 * round(tenure / 7), y = friendships_initiated / tenure),
       data = subset(pf, tenure > 0)) +
  geom_line(aes(color = year_joined.bucket),
            stat = "summary",
            fun.y = mean)

ggplot(aes(x = 30 * round(tenure / 30), y = friendships_initiated / tenure),
       data = subset(pf, tenure > 0)) +
  geom_line(aes(color = year_joined.bucket),
            stat = "summary",
            fun.y = mean)

ggplot(aes(x = 90 * round(tenure / 90), y = friendships_initiated / tenure),
       data = subset(pf, tenure > 0)) +
  geom_line(aes(color = year_joined.bucket),
            stat = "summary",
            fun.y = mean)

ggplot(aes(x = 7 * round(tenure / 7), y = friendships_initiated / tenure),
       data = subset(pf, tenure > 0)) +
  geom_smooth(aes(color = year_joined.bucket))


```

***

### Sean's NFL Fan Sentiment Study
Notes:

***

### Introducing the Yogurt Data Set
Notes:

***

### Histograms Revisited
Notes:

```{r Histograms Revisited}
#7.18
yo <- read.csv('yogurt.csv')

str(yo)
yo$id <- factor(yo$id)

ggplot(aes(x=price), data=yo) +
  geom_histogram(binwidth = 10)

```

***

### Number of Purchases
Notes:

```{r Number of Purchases}
#7.19

yo$all.purchases <- yo$strawberry + yo$blueberry + yo$pina.colada + yo$plain + yo$mixed.berry

#using transform:
yo <- transform(yo, all.purchases = strawberry + blueberry + pina.colada + plain + mixed.berry)

str(yo)
summary(yo$all.purchases)

ggplot(aes(x=all.purchases), data=yo) + 
  geom_histogram(binwidth = 1)

```

***

### Prices over Time
Notes:

```{r Prices over Time}

#7.20
ggplot(aes(x=time, y=price), data = yo) +
  geom_point(alpha=1/10)

#answer solution
ggplot(aes(x=time, y=price), data = yo) +
  geom_jitter(alpha=1/4, shape = 21, fill=I('#F79420'))


```

***

### Sampling Observations
Notes:

***

### Looking at Samples of Households

```{r Looking at Sample of Households}
#7.22
set.seed(4230)
sample.ids <- sample(levels(yo$id), 16)

ggplot(aes(x=time, y=price), data = subset(yo, id %in% sample.ids)) +
  geom_point(aes(size = all.purchases),pch =1)
 
#enhanced
ggplot(aes(x=time, y=price), data = subset(yo, id %in% sample.ids)) +
  facet_wrap( ~ id ) +
  geom_line() +
  geom_point(aes(size = all.purchases),pch =1)

#my version
set.seed(2650)
sample.ids <- sample(levels(yo$id), 16)

ggplot(aes(x=time, y=price), data = subset(yo, id %in% sample.ids)) +
  facet_wrap( ~ id ) +
  geom_line() +
  geom_point(aes(size = all.purchases),pch =1)

         


```

***

### The Limits of Cross Sectional Data
Notes:

***

### Many Variables
Notes:

***

### Scatterplot Matrix
Notes:

```{r}
library(GGally)

theme_set(theme_minimal(20))
set.seed(1836)
pf_subset <- pf[, c(2:15)]
names(pf_subset)
ggpairs(pf_subset[sample.int(nrow(pf_subset), 1000),])

```

***

### Even More Variables
Notes:

***

### Heat Maps
Notes:

```{r}
nci <- read.table("nci.tsv")
colnames(nci) <- c(1:64)
head(nci)
```

```{r}
nci.long.samp <- melt(as.matrix(nci[1:200,]))
names(nci.long.samp) <- c("gene", "case", "value")
head(nci.long.samp)

ggplot(aes(y = gene, x = case, fill = value),
  data = nci.long.samp) +
  geom_tile() +
  scale_fill_gradientn(colours = colorRampPalette(c("blue", "red"))(100))
```


***

### Analyzing Three of More Variables
Reflection:

***

Click **KnitHTML** to see all of your hard work and to have an html
page of this lesson, your answers, and your notes!

