---
output: html_document
editor_options: 
  chunk_output_type: console
---
Lesson 3
========================================================

***

### What to Do First?
Notes: Open data set

***
```{r}
knitr::opts_chunk$set(echo = TRUE)
```

### Pseudo-Facebook User Data
Notes:

```{r Pseudo-Facebook User Data}
list.files()
pf <- read.delim('pseudo_facebook.tsv')
head(pf)
names(pf)
```

***

### Histogram of Users' Birthdays
Notes:

```{r Histogram of Users\' Birthdays}
#install.packages('ggplot2')
library(ggplot2)
qplot(x=dob_day, data=pf) + scale_x_continuous(breaks=1:31)
```
```{r}
library(ggthemes)
theme_set(theme_minimal(24)) 

qplot(x=dob_day, data=pf) +
  scale_x_continuous(breaks=1:31) +
  facet_wrap(~dob_month, ncol = 3)



```
  
***

#### What are some things that you notice about this histogram?
Response:

***

### Moira's Investigation
Notes:

***
#Plot with qplot
```{r}
qplot(x=friend_count, data=pf, xlim=c(0,1000))
```

#plot with ggplot
```{r}
ggplot(aes(x = friend_count), data = pf) +
  geom_histogram()
```

#plot with qplot, modified x axis to cap at 1000 and put in 50 point breaks (scale x), modified binwidth, 
```{r}
qplot(x=friend_count, data=subset(pf, !is.na(gender)), bindwidth=25) +
  scale_x_continuous(limits = c(0,1000), breaks=seq(0,1000,50)) +
  facet_grid(~gender)

#NOTE: could also use na.omit(pf) but it will remove ALL NA values, not just for a single column
```

#plot with ggplot, modified x axis
```{r}
ggplot(aes(x = friend_count), data = pf) +
  geom_histogram() +
  scale_x_continuous(limits = c(0, 1000), breaks = seq(0, 1000, 50)) +
  facet_wrap(~gender)
```

```{r}
qplot(x = friend_count, data = pf) +
  facet_grid(gender ~ .)
```

```{r}
table(pf$gender)
by(pf$friend_count, pf$gender, summary)
```


### Estimating Your Audience Size
Notes:

***

#### Think about a time when you posted a specific message or shared a photo on Facebook. What was it?
Response:

#### How many of your friends do you think saw that post?
Response:

#### Think about what percent of your friends on Facebook see any posts or comments that you make in a month. What percent do you think that is?
Response:

***

### Perceived Audience Size
Notes:

***
### Faceting
Notes:

```{r Faceting}

```

#### Let’s take another look at our plot. What stands out to you here?
Response:

***

### Be Skeptical - Outliers and Anomalies
Notes:

***

### Moira's Outlier
Notes:
#### Which case do you think applies to Moira’s outlier?
Response:

***

### Friend Count
Notes:

#### What code would you enter to create a histogram of friend counts?

```{r Friend Count}

```

#### How is this plot similar to Moira's first plot?
Response:

***

### Limiting the Axes
Notes:

```{r Limiting the Axes}

```

### Exploring with Bin Width
Notes:

***

### Adjusting the Bin Width
Notes:

### Faceting Friend Count
```{r Faceting Friend Count}
# What code would you add to create a facet the histogram by gender?
# Add it to the code below.
qplot(x = friend_count, data = pf, binwidth = 10) +
  scale_x_continuous(limits = c(0, 1000),
                     breaks = seq(0, 1000, 50))
```

***

### Omitting NA Values
Notes:

```{r Omitting NA Values}

```

***

### Statistics 'by' Gender
Notes:

```{r Statistics \'by\' Gender}

```

#### Who on average has more friends: men or women?
Response:

#### What's the difference between the median friend count for women and men?
Response:

#### Why would the median be a better measure than the mean?
Response:

***

### Tenure
Notes:

```{r Tenure}
qplot(x=tenure, data=pf, binwidth=30,
      color = I('black'), fill=I('#099DD9'))
```

***

#### How would you create a histogram of tenure by year?

```{r Tenure Histogram by Year}
qplot(x=(pf$tenure/365), data=pf, binwidth=.25,
      color = I('black'), fill=I('#099DD9')) +
  scale_x_continuous(limits = c(0,7), breaks=seq(0,7,.5))

#qplot(x=friend_count, data=subset(pf, !is.na(gender)), bindwidth=25) +
#  scale_x_continuous(limits = c(0,1000), breaks=seq(0,1000,50)) +

```


***

### Labeling Plots
Notes:

```{r Labeling Plots}
qplot(x=tenure/365, data=pf, binwidth=.25,
      color = I('black'), fill=I('#099DD9'),
      xlab='Number of Years using Facebook',
      ylab='Number of Users in Sample') +
  scale_x_continuous(limits = c(0,7), breaks=seq(0,7,.5))
```

***

### User Ages
Notes:

```{r User Ages}
qplot(x=age, data=pf, binwidth=1,
      color = I('black'), fill=I('#099DD9'),
      xlab='Age of Facebook Users',
      ylab='Number of Users in Sample') +
  scale_x_continuous(limits = c(10,120), breaks=seq(10,120,10))

```

#### What do you notice?
Response:

***

### The Spread of Memes
Notes:

***

### Lada's Money Bag Meme
Notes:

***

### Transforming Data
Notes:

***

### Add a Scaling Layer
Notes:
```{r}
#install.packages('gridExtra')
library(gridExtra)
```

```{r Add a Scaling Layer}
qplot(x=friend_count, data=pf)

summary(pf$friend_count)

summary(log10(pf$friend_count+1))

summary(sqrt(pf$friend_count))
```

```{r}
p1 <- qplot(x=friend_count, data=pf)
p2 <- qplot(x=log10(friend_count+1), data=pf)
p3 <- qplot(x=sqrt(friend_count), data= pf)

grid.arrange(p1,p2,p3, ncol=1)
```

```{r}
#same as above done with ggplot

p1 <- ggplot(aes(x=friend_count), data=pf) + geom_histogram()
p2 <- p1 + scale_x_log10()
p3 <- p1 + scale_x_sqrt()

grid.arrange(p1,p2,p3, ncol=1)

```

```{r}
logScale <- qplot(x=log10(friend_count), data =pf)

countScale <- ggplot(aes(x=friend_count), data=pf) +
  geom_histogram() +
  scale_x_log10()

grid.arrange(logScale, countScale,ncol = 2)

#can add scale_x_log10 to any plot
qplot(x=friend_count, data=pf) +
  scale_x_log10()


```


***


### Frequency Polygons

```{r Frequency Polygons}
qplot(x=friend_count, y=..count../sum(..count..),
      data=subset(pf, !is.na(gender)), bindwidth=10,
      ylab='Proportion of users with that friend count',
      xlab='Friend Count',
      geom='freqpoly', color = gender) +
  scale_x_continuous(limits = c(0,1000), breaks=seq(0,1000,50))
  


```

***

### Likes on the Web
Notes:

```{r Likes on the Web}
qplot(x=www_likes, y=..count../sum(..count..),
      data=subset(pf, !is.na(gender)),
      ylab='Proportion of users that make likes',
      xlab='Like Count',
      geom='freqpoly', color = gender) +
  scale_x_continuous() +
  scale_x_log10()

summary(pf$www_likes)


by(pf$www_likes, pf$gender, sum)
```


***

### Box Plots
Notes:

```{r Box Plots}
qplot(x=gender, y=friend_count, 
      data=subset(pf, !is.na(gender)),
      geom='boxplot', ylim=c(0,1000)) 
# can you ylim or this: +  scale_y_continuous(limits=c(0,1000))

qplot(x=gender, y=friend_count, 
      data=subset(pf, !is.na(gender)),
      geom='boxplot') +
  coord_cartesian(ylim=c(0,1000))
#using coord_cartesian assures that the data summary below will match the box plot(Data will not be removed with this paramater)

by(pf$friend_count, pf$gender, summary)

```

#### Adjust the code to focus on users who have friend counts between 0 and 1000.

```{r}

```

***

### Box Plots, Quartiles, and Friendships
Notes:

```{r Box Plots, Quartiles, and Friendships}

```

#### On average, who initiated more friendships in our sample: men or women?
Response:
#### Write about some ways that you can verify your answer.
Response:
```{r Friend Requests by Gender}
names(pf)
summary(pf$friendships_initiated)
by(pf$friendships_initiated, pf$gender, summary)

qplot(x=gender, y=friendships_initiated, 
      data=subset(pf, !is.na(gender)),
      geom='boxplot') +
  coord_cartesian(ylim=c(0,500))

by(pf$friendships_initiated, pf$gender, sum)
```

Response:

***

### Getting Logical
Notes:

```{r Getting Logical}
summary(pf$mobile_likes)
summary(pf$mobile_likes > 0)

mobile_checkin <- NA
pf$mobile_checkin <- ifelse(pf$mobile_likes > 0, 1, 0)
pf$mobile_checkin <- factor(pf$mobile_checkin)

summary(pf$mobile_checkin)

sum(pf$mobile_checkin == 1) / (sum(pf$mobile_checkin == 0) + sum(pf$mobile_checkin == 1))

# - or -

sum(pf$mobile_checkin == 1) / length(pf$mobile_checkin)



```

Response:

***

### Analyzing One Variable
Reflection:

***

Click **KnitHTML** to see all of your hard work and to have an html
page of this lesson, your answers, and your notes!