---
output: html_document
editor_options: 
  chunk_output_type: console
---
Lesson 6
========================================================

### Welcome
Notes:

***

### Scatterplot Review

```{r Scatterplot Review}
library(ggplot2)
data(diamonds)

ggplot(data=diamonds, aes(y=diamonds$price, x=diamonds$carat)) +
  geom_point() +
  xlim(NA,quantile(diamonds$carat,.99)) +
  ylim(NA,quantile(diamonds$price,.99))

#sexy  version
ggplot(data=diamonds, aes(y=diamonds$price, x=diamonds$carat)) +
  geom_point(fill=I('#F79420'), color=I('black'), shape =22) +
  xlim(0,quantile(diamonds$carat,.99)) +
  ylim(0,quantile(diamonds$price,.99))


#9.4 
ggplot(data=diamonds, aes(y=diamonds$price, x=diamonds$carat)) +
  geom_point(fill=I('#F79420'),alpha =0.25) +
  xlim(0,quantile(diamonds$carat,.99)) +
  ylim(0,quantile(diamonds$price,.99)) +
  stat_smooth(method='lm')


```

***

### Price and Carat Relationship
Response:

***

### Frances Gerety
Notes:

#### A diamonds is


***

### The Rise of Diamonds
Notes:

***

### ggpairs Function
Notes:

```{r ggpairs Function}
# install these if necessary
#install.packages('GGally')
#install.packages('scales')
install.packages('memisc')
install.packages('lattice')
install.packages('MASS')
#install.packages('car')
#install.packages('reshape')
#install.packages('plyr')

# load the ggplot graphics package and the others
library(ggplot2)
library(GGally)
library(scales)
library(memisc)

# sample 10,000 diamonds from the data set
set.seed(20022012)
diamond_samp <- diamonds[sample(1:length(diamonds$price), 10000), ]
ggpairs(diamond_samp, 
        lower = list(continuous = wrap("points", shape = I('.'))), 
        upper = list(combo = wrap("box", outlier.shape = I('.'))))
```

What are some things you notice in the ggpairs output?
Response:

***

### The Demand of Diamonds
Notes:

```{r The Demand of Diamonds}
#9.8
# Create two histograms of the price variable
# and place them side by side on one output image.

# We've put some code below to get you started.

# The first plot should be a histogram of price
# and the second plot should transform
# the price variable using log10.

# Set appropriate bin widths for each plot.
# ggtitle() will add a title to each histogram.

# You can self-assess your work with the plots
# in the solution video.

# ALTER THE CODE BELOW THIS LINE
# ==============================================

library(gridExtra)

plot1 <- qplot(x=price, data=diamonds, binwidth=100) +
  ggtitle('Price')

plot2 <- qplot(x=price, data=diamonds, binwidth=.01) +
  ggtitle('Price (log10)') +
  scale_x_log10()

grid.arrange(plot1,plot2,ncol=1)


```

***

### Connecting Demand and Price Distributions
Notes:

***

### Scatterplot Transformation

```{r Scatterplot Transformation}
#9.10

qplot(carat, price, data=diamonds) +
  scale_y_continuous(trans=log10_trans()) +
  ggtitle('Price (log10) by Carat')

```


### Create a new function to transform the carat variable

```{r cuberoot transformation}
cuberoot_trans = function() trans_new('cuberoot', transform = function(x) x^(1/3),
                                      inverse = function(x) x^3)
```

#### Use the cuberoot_trans function
```{r Use cuberoot_trans}
ggplot(aes(carat, price), data = diamonds) + 
  geom_point() + 
  scale_x_continuous(trans = cuberoot_trans(), limits = c(0.2, 3),
                     breaks = c(0.2, 0.5, 1, 2, 3)) + 
  scale_y_continuous(trans = log10_trans(), limits = c(350, 15000),
                     breaks = c(350, 1000, 5000, 10000, 15000)) +
  ggtitle('Price (log10) by Cube-Root of Carat')
```

***

### Overplotting Revisited

```{r Sort and Head Tables}
#9.11
head(sort(table(diamonds$carat), decreasing = T))
head(sort(table(diamonds$price), decreasing = T))

```


```{r Overplotting Revisited}
# Add a layer to adjust the features of the
# scatterplot. Set the transparency to one half,
# the size to three-fourths, and jitter the points.

# If you need hints, see the Instructor Notes.
# There are three hints so scroll down slowly if
# you don't want all the hints at once.

# ALTER THE CODE BELOW THIS LINE
# =======================================================================

ggplot(aes(carat, price), data = diamonds) + 
  geom_point(alpha=.5, size =.75, position='jitter') + 
  scale_x_continuous(trans = cuberoot_trans(), limits = c(0.2, 3),
                     breaks = c(0.2, 0.5, 1, 2, 3)) + 
  scale_y_continuous(trans = log10_trans(), limits = c(350, 15000),
                     breaks = c(350, 1000, 5000, 10000, 15000)) +
  ggtitle('Price (log10) by Cube-Root of Carat')
```

***

### Other Qualitative Factors
Notes:

***

### Price vs. Carat and Clarity

Alter the code below.
```{r Price vs. Carat and Clarity}

# Adjust the code below to color the points by clarity.

# A layer called scale_color_brewer() has 
# been added to adjust the legend and
# provide custom colors.

# See if you can figure out what it does.
# Links to resources are in the Instructor Notes.

# You will need to install the package RColorBrewer
# in R to get the same colors and color palettes.

# =======================================

# install and load the RColorBrewer package
?diamonds
install.packages('RColorBrewer')
library(RColorBrewer)

ggplot(aes(x = carat, y = price, color=clarity), data = diamonds) + 
  geom_point(alpha = 0.5, size = 1, position = 'jitter') +
  scale_color_brewer(type = 'div',
    guide = guide_legend(title = 'Clarity', reverse = T,
    override.aes = list(alpha = 1, size = 2))) +  
  scale_x_continuous(trans = cuberoot_trans(), limits = c(0.2, 3),
    breaks = c(0.2, 0.5, 1, 2, 3)) + 
  scale_y_continuous(trans = log10_trans(), limits = c(350, 15000),
    breaks = c(350, 1000, 5000, 10000, 15000)) +
  ggtitle('Price (log10) by Cube-Root of Carat and Clarity')
```

***

### Clarity and Price
Response:

***

### Price vs. Carat and Cut

Alter the code below.
```{r Price vs. Carat and Cut}
#9.15
# Let's look at cut and see if we find a similar result.

# Adjust the code below to color the points by cut.
# Change any other parts of the code as needed.

ggplot(aes(x = carat, y = price, color = cut), data = diamonds) + 
  geom_point(alpha = 0.5, size = 1, position = 'jitter') +
  scale_color_brewer(type = 'div',
                     guide = guide_legend(title = 'Cut', reverse = T,
                                          override.aes = list(alpha = 1, size = 2))) +  
  scale_x_continuous(trans = cuberoot_trans(), limits = c(0.2, 3),
                     breaks = c(0.2, 0.5, 1, 2, 3)) + 
  scale_y_continuous(trans = log10_trans(), limits = c(350, 15000),
                     breaks = c(350, 1000, 5000, 10000, 15000)) +
  ggtitle('Price (log10) by Cube-Root of Carat and Cut')
```

***

### Cut and Price
Response:

***

### Price vs. Carat and Color

Alter the code below.
```{r Price vs. Carat and Color}
#9.17
# Finally, let's use diamond color to color our plot.

# Adjust the code below to color the points by diamond colors
# and change the titles.


#DROP reverse=T to reverse the order of the legend to match the data itself
ggplot(aes(x = carat, y = price, color = color), data = diamonds) + 
  geom_point(alpha = 0.5, size = 1, position = 'jitter') +
  scale_color_brewer(type = 'div',
                     guide = guide_legend(title = 'Color', reverse = T,
                                          override.aes = list(alpha = 1, size = 2))) +  
  scale_x_continuous(trans = cuberoot_trans(), limits = c(0.2, 3),
                     breaks = c(0.2, 0.5, 1, 2, 3)) + 
  scale_y_continuous(trans = log10_trans(), limits = c(350, 15000),
                     breaks = c(350, 1000, 5000, 10000, 15000)) +
  ggtitle('Price (log10) by Cube-Root of Carat and Color')
```

***

### Color and Price
Response:

***

### Linear Models in R
Notes:

Response:

***

### Building the Linear Model
Notes:

```{r Building the Linear Model}

#"I" stands for "as is"

m1 <- lm(I(log(price)) ~ I(carat^(1/3)), data = diamonds)
m2 <- update(m1, ~ . + carat)
m3 <- update(m2, ~ . + cut)
m4 <- update(m3, ~ . + color)
m5 <- update(m4, ~ . + clarity)
mtable(m1, m2, m3, m4, m5)
  ```

Notice how adding cut to our model does not help explain much of the variance
in the price of diamonds. This fits with out exploration earlier.

***

### Model Problems
Video Notes:

Research:
(Take some time to come up with 2-4 problems for the model)
(You should 10-20 min on this)

Response:
Global economic factors
Inflation
Year of the data (2008-1014) - how much has changed since then
Geo-polical factors
The increasing age at which people get married (older people = higher incomes, willingness to pay more)

***

### A Bigger, Better Data Set
Notes:

```{r A Bigger, Better Data Set}
install.package('bitops')
install.packages('RCurl')
library('bitops')
library('RCurl')

diamondsurl = getBinaryURL("https://raw.github.com/solomonm/diamonds-data/master/BigDiamonds.Rda")
load("BigDiamonds.rda")

```

The code used to obtain the data is available here:
https://github.com/solomonm/diamonds-data

## Building a Model Using the Big Diamonds Data Set
Notes:

```{r Building a Model Using the Big Diamonds Data Set}
m1 <- lm(I(log(price)) ~ I(carat^(1/3)),
         data=subset(diamondsbig, price < 10000 & cert == "GIA"))
m2 <- update(m1, ~ . + carat)
m3 <- update(m2, ~ . + cut)
m4 <- update(m3, ~ . + color)
m5 <- update(m4, ~ . + clarity)
mtable(m1, m2, m3, m4, m5)

#Instructor answer
diamondsbig$logprice = log(diamondsbig$price)
m1 <- lm(logprice ~ I(carat^(1/3)),
         data=diamondsbig[diamondsbig$price < 10000 & diamondsbig$cert == "GIA",])
m2 <- update(m1, ~ . + carat)
m3 <- update(m2, ~ . + cut)
m4 <- update(m3, ~ . + color)
m5 <- update(m4, ~ . + clarity)
mtable(m1, m2, m3, m4, m5)


  ```

```


***

## Predictions

Example Diamond from BlueNile:
Round 1.00 Very Good I VS1 $5,601

```{r}
#Be sure you’ve loaded the library memisc and have m5 saved as an object in your workspace.
library(memisc)

thisDiamond = data.frame(carat = 1.00, cut = "V.Good",
                         color = "I", clarity="VS1")
modelEstimate = predict(m5, newdata = thisDiamond,
                        interval="prediction", level = .95)

#since we took log of price, we need to exponentiate the answer
exp(modelEstimate)



dat = data.frame(m4$model, m4$residuals)

with(dat, sd(m4.residuals))

with(subset(dat, carat > .9 & carat < 1.1), sd(m4.residuals))

dat$resid <- as.numeric(dat$m4.residuals)
ggplot(aes(y = resid, x = round(carat, 2)), data = dat) +
  geom_line(stat = "summary", fun.y = sd)



```

Evaluate how well the model predicts the BlueNile diamond's price. Think about the fitted point estimate as well as the 95% CI.

***

## Final Thoughts
Notes:

***

Click **KnitHTML** to see all of your hard work and to have an html
page of this lesson, your answers, and your notes!

